Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils

### 2020-10-26 doc/manuscript.Rmd
corrected typos

It is evident that the groundwater buffer effect will play a determinant role in the response of mire habitats to climate change [@RN4936]. Both bogs and fens are actively peat forming and depend on precipitation, but fens rely also on sources of telluric water from mineral ground. Moreover, they show a mosaic of microreliefs and a specific plant cover zonation which makes them especially sensitive to small changes in wetness. Climate change affects the quantity, timing and spatial distribution of precipitation, leading to changes in surface wetness which alter the intensity or organic decomposition by disturbing the conditions for plant grow and the depth of air penetration [@RN4935]. Global warming would also result in warmer groundwater delivered to fens, but there is a gap of knowledge about the rate of such changes and their ecological consequences within the mires [@RN4936].

@@

One important difference with previous studies is the magnitude of the buffer during winter. The articles that had used model air temperatures as a control concluded that the buffer effect was stronger at the cold end of the thermal spectrum, i.e. in winter and at night [@RN2356; @RN4675]. In the case of this investigation, the situation is the reverse: the effect is weaker when considering the minimum temperatures of the cold period. This indicates the importance of identifying root temperatures when working with plant communities. At high elevations of the study region, the soil can remain covered by snow for periods of winter, and this has its own insulating effect on soil temperatures [@RN2392]. Indeed, snow cover has being described as one of the vertical features that affects vegetation distribution in a local manner [@RN4760]. In any case, since this study was conducted at the southern limit of mire distribution, more research on the winter buffer is needed at high latitudes, since it might have a deeper biological significance there.

In these southern mires, the buffering effect is much stronger during the summer (**Fig. 5**). This is most noticeable in the two lowest elevations, El Molinucu and La Malva. La Malva shows especially high summer temperatures within the dry soils (> 40 ºC). It must be noted that this difference of > 15 ºC between the dry and waterlogged points of La Malva, which are separated by a few meters, is almost four times the warming expected in the study region at the end of this century (+ 4 ºC) under the more extreme emissions scenario [@RN4939]. Whereas the rest of the study sites are flat, La Malva is a calcareous spring in a southeast-facing slope on limestone, a place experiencing sub-Mediterranean conditions at the micro-scale [@RN4681], and surrounded by a forest of evergreen oaks (*Quercus rotundifolia* Lam., *Quercus faginea* Lam.). This suggests the importance of groundwater in the existence of mire vegetation in Mediterranean areas [@RN4682], not only from the water-availability aspect, but also providing cooling regulation during summer [@RN3344]. The buffering is also prominent at one of the highest elevations, La Recoleta, which may be explained by the high saturation of water in this kind of habitat, a quaking bog, i.e. a former glacial lake undergoing silting. However, the relatively small number of sites in our study does not allow to explore further the moderating factors that may drive differences in the strength of the buffer. More studies focusing on each kind of habitat are needed to understand the relation between hydrology conditions and microclimatic effects.

@@

E.F.P. received financial support from the Government of Asturias and the FP7 – Marie Curie - COFUND programme of the European Commission (Grant ‘Clarín’ ACB17-19).

>>>>

It is evident that the groundwater buffer effect will play a determinant role in the response of mire habitats to climate change [@RN4936]. Both bogs and fens are actively peat forming and depend on precipitation, but fens rely also on sources of telluric water from mineral ground. Moreover, they show a mosaic of microreliefs and a specific plant cover zonation which makes them especially sensitive to small changes in wetness. Climate change affects the quantity, timing and spatial distribution of precipitation, leading to changes in surface wetness which alter the intensity of organic decomposition by disturbing the conditions for plant grow and the depth of air penetration [@RN4935]. Global warming would also result in warmer groundwater delivered to fens, but there is a gap of knowledge about the rate of such changes and their ecological consequences within the mires [@RN4936].

@@

One important difference with previous studies is the magnitude of the buffer during winter. The articles that had used model air temperatures as a control concluded that the buffer effect was stronger at the cold end of the thermal spectrum, i.e. in winter and at night [@RN2356; @RN4675]. In the case of this investigation, the situation is the reverse: the effect is weaker when considering the minimum temperatures of the cold period. This indicates the importance of identifying root temperatures when working with plant communities. At high elevations of the study region, the soil can remain covered by snow for periods of winter, and this has its own insulating effect on soil temperatures [@RN2392]. Indeed, snow cover has been described as one of the vertical features that affects vegetation distribution in a local manner [@RN4760]. In any case, since this study was conducted at the southern limit of mire distribution, more research on the winter buffer is needed at high latitudes, since it might have a deeper biological significance there.

In these southern mires, the buffering effect is much stronger during the summer (**Fig. 5**). This is most noticeable in the two lowest elevations, El Molinucu and La Malva. La Malva shows especially high summer temperatures within the dry soils (> 40 ºC). It must be noted that this difference of > 15 ºC between the dry and waterlogged points of La Malva, which are separated by a few meters, is almost four times the warming expected in the study region at the end of this century (+ 4 ºC) under the most extreme emissions scenario [@RN4939]. Whereas the rest of the study sites are flat, La Malva is a calcareous spring in a southeast-facing slope on limestone, a place experiencing sub-Mediterranean conditions at the micro-scale [@RN4681], and surrounded by a forest of evergreen oaks (*Quercus rotundifolia* Lam., *Quercus faginea* Lam.). This suggests the importance of groundwater in the existence of mire vegetation in Mediterranean areas [@RN4682], not only from the water-availability aspect, but also providing cooling regulation during summer [@RN3344]. The buffering is also prominent at one of the highest elevations, La Recoleta, which may be explained by the high saturation of water in this kind of habitat, a quaking bog, i.e. a former glacial lake undergoing silting. However, the relatively small number of sites in our study does not allow to explore further the moderating factors that may drive differences in the strength of the buffer. More studies focusing on each kind of habitat are needed to understand the relation between hydrology conditions and microclimatic effects.

@@

Eduardo Fernández-Pascual received financial support from the Government of Asturias and the FP7 – Marie Curie – COFUND programme of the European Commission (Grants “Clarín” ACA14-19 and ACB17-19). Logger installation was partly supported by the LIFE+ programme of the European Commission (project LIFE11/NAT/ES/707 - TREMEDAL).
